Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2021-10-11 analysis/paper/paper.Rmd
AGH second revisions to first draft

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author:
  - Bjørn Peare Bartholdy
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    #template: "../templates/sn-article-template/sn-article.tex"
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
    extra_dependencies: "subfig"
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
  bookdown::html_document2:
    df_print: kable
    keep_md: true
linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus has proven to contain a wealth of information 
 on the dietary habits of past populations. These insights have, to a large extent,
 been obtained by the extraction and identification of
 starch granules contained within the mineralised dental plaque from a wide range 
 of regions and time periods. The scope of previous studies have been limited to 
 microfossil extraction and identification to reconstruct dietary preferences from
 the archaeological record, and few studies have attempted to address the biases 
 of starch retention in dental calculus. Those that have considered this problem 
 have been limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental 
 research on  starch incorporation and retention to be conducted on *in vitro* 
 dental calculus in a controlled laboratory setting. The biofilms were exposed 
 to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
 and counted. We show that the granule counts extracted from the model dental calculus 
 represented a low proportion (ranging from 0.064% to 0.161%) of the total number 
 of granules exposed to the biofilms throughout the experiment.
 Additionally, we found that the ratios of granule sizes from the extracted 
 starch granules differed from the original treatment solutions, with large 
 granules (>20 $\mu$m) consistently being under-represented.
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.30, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
#cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
## end refs.bib ##

@@

Word count: `r wd_count`

@@

All scripts and data used in the analysis are available on OSF 
(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch).
More detailed protocols are available on protocols.io.

@@

<!--

```{r amylase-activity-tbl}
rownames(plt1_ph1_result) <- rows
colnames(plt1_ph1_result) <- cols
knitr::kable(signif(plt1_ph1_result, 3), row.names = TRUE, caption = "Amylase activity in U/mL enzyme, where a U is mg maltose released from starch in six minutes at 36 degrees Celsius. Plate 1, photometric reading 1. Negative values converted to 0.")
```
-->

@@

<!-- INSERT MICROSCOPE PICTURES OF STARCHES -->

@@

# count_samp_comb_per <- count_sample_comb %>%
#   mutate(across(c(s, m, l, total), 
#                 function(x) 
#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))

# count_samp_comb_out <- tibble(count_samp_comb_per, count_sample_comb_sd, .name_repair = "unique") %>%
#   select(1,2,3,7,4,8,5,9,6,10) #%>%
  #add_row(tibble_row(count_sample_comb_out), .before = 3)

@@

```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

<!--
```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```
 -->

@@

```{r cor, echo=FALSE}
# prepare cor output for reporting
cor_r <- signif(count_cor$estimate, 3)
cor_ci <- signif(count_cor$conf.int, 3)
cor_p <- if(count_cor$p.value > 0.001){
  signif(count_cor$p.value, 3)
} else {
  paste("< 0.001")
}

cor2_r <- signif(starch_cor$estimate, 3)
cor2_ci <- signif(starch_cor$conf.int, 3)
cor2_p <- if(starch_cor$p.value > 0.001){
  signif(starch_cor$p.value, 3)
} else {
  paste("< 0.001")
}
# Convert r correlation result to text
direct_cor <- ifelse(cor_r < 0, "negative", "positive")
strength_cor <- if(cor_r >= 0.8){
  "very strong"
} else if(cor_r < 0.8 & cor_r >= 0.6){
  "strong"
} else if(cor_r < 0.6 & cor_r >= 0.4){
  "moderate"
} else if(cor_r < 0.4 & cor_r >= 0.2){
  "weak"
} else{
  "very weak"
}
```

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

Pearson's *r* suggests a 
`r strength_cor` `r direct_cor` 
correlation between the total weight of the biofilms and the total starch count 
(standardised by z-score) extracted from the samples across treatments, 
*r* = `r cor_r`, 
90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
p = `r cor_p`
(Figure \@ref(fig:cor-plot)).

@@

<!-- size ratios in the larger deposits vs the smaller deposits?
  Are larger starches more proportionately represented in larger deposits?
  -->

```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
pl_cor2
```

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot));
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2)).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
Further research is needed to determine why this differs from previous archaeological
findings.
<!--
Given the low proportion of starch granule retention in our dental calculus model,
and the limited difference between treatments, we can conjecture that starch 
availability is not a main factor in starch incorporation. While there was a
proportional incorporation, i.e. more granules in the solution led to more granules
in the calculus, the difference between solutions [a number/percentage] was of a
higher magnitude than the difference between samples <!-- a number/percentage-->

@@

<!-- add potential for morphological differences affecting incorporation 
 Medium potato starches had the highest rate of incorporation 
 All large starches had a consistently low rate of incorporation 
 Starch granule adhesion to bacteria and enzymes? Do they have a preference?
 Binding to select species in a biofilm could explain clustering
 Colonisation of bacteria onto starch @nagaraBacteriaNutrients2017 (although not
 clear if the bacteria in our model are even 'consuming' starch-->

<!-- Combined, the effect of intra-oral starch incorporation -->
<!-- and the bias against large granules, together with the increased effect of taphonomy -->
<!-- on  small granules, would eliminate a large portion of the consumed starch granules, -->
<!-- and could explain the large descrepency between our counts (unaffected by hydrolysis  -->
<!-- and taphonomy) and the aforementioned studies.  -->

@@

<!-- If you're reading this, thank you for taking the time to read the source 
documents. You're awesome! -->
This research has received funding from the European Research Council under the 
European Union’s Horizon 2020 research and innovation program, grant agreement 
number STG–677576 (“HARVEST”). We also thank everyone in the general vicinity
of the lab for enduring the smell of bacterial accumulation.
We did NOT make use of Sci-Hub to access articles stuck behind a paywall, 
and will, therefore, NOT acknowledge the use of Sci-Hub in this study.

>>>>

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author: |
 Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
date: |
 $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
output:
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::pdf_document2:
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      #before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
    #extra_dependencies: "subfig"
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
#linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus has proven to contain a wealth of information 
 on the dietary habits of past populations. These insights have, to a large extent,
 been obtained by the extraction and identification of
 starch granules contained within the mineralised dental plaque from a wide range 
 of regions and time periods. The scope of previous studies have been limited to 
 microfossil extraction and identification to reconstruct dietary preferences from
 the archaeological record, and few studies have attempted to address the biases 
 of starch retention in dental calculus. Those that have considered this problem 
 have been limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental 
 research on  starch incorporation and retention to be conducted on *in vitro* 
 dental calculus in a controlled laboratory setting. The biofilms were exposed 
 to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
 and counted. We show that the granule counts extracted from the model dental calculus 
 represented a low proportion (ranging from 0.064% to 0.161%) of the total number 
 of granules exposed to the biofilms throughout the experiment.
 Additionally, we found that the ratios of granule sizes from the extracted 
 starch granules differed from the original treatment solutions, with large 
 granules (>20 $\mu$m) consistently being under-represented.
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.30, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/",
  fig.align = "center")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
#cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
#cite_keys <- unlist(cite_keys, use.names = FALSE)
#rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
## end refs.bib ##

@@

@@

@@

@@

@@

@@

```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.", out.width="80%"}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

```{r cor, echo=FALSE}
# prepare cor output for reporting
cor_r <- signif(count_cor$estimate, 3)
cor_ci <- signif(count_cor$conf.int, 3)
cor_p <- if(count_cor$p.value > 0.001){
  paste("=", signif(count_cor$p.value, 3))
} else {
  paste("< 0.001")
}

cor2_r <- signif(starch_cor$estimate, 3)
cor2_ci <- signif(starch_cor$conf.int, 3)
cor2_p <- if(starch_cor$p.value > 0.001){
  paste("=", signif(starch_cor$p.value, 3))
} else {
  paste("< 0.001")
}
# Convert r correlation result to text
direct_cor <- ifelse(cor_r < 0, "negative", "positive")
strength_cor <- if(cor_r >= 0.8){
  "very strong"
} else if(cor_r < 0.8 & cor_r >= 0.6){
  "strong"
} else if(cor_r < 0.6 & cor_r >= 0.4){
  "moderate"
} else if(cor_r < 0.4 & cor_r >= 0.2){
  "weak"
} else{
  "very weak"
}
```

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.", out.width="80%"}
pl_cor
#pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

Pearson's *r* suggests a 
`r strength_cor` `r direct_cor` 
correlation between the total weight of the biofilms and the total starch count 
(standardised by z-score) extracted from the samples across treatments, 
*r* = `r cor_r`, 
90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
p `r cor_p`
(Figure \@ref(fig:cor-plot)).

```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.", out.width="80%"}
pl_cor2
```

@@

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot));
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2)).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
Further research is needed to determine why this differs from previous archaeological
findings.

@@

@@

<!-- If you're reading this, thank you for taking the time to read the source 
documents. You're awesome! -->
We would like to thank Dr. Stephanie Schnorr for help with the amylase activity protocol.
We also thank everyone in the general vicinity
of the lab for enduring the smell of bacterial accumulation.
Since we did NOT make use of Sci-Hub to access articles stuck behind a paywall, 
we will NOT acknowledge the use of Sci-Hub in this study.

This research has received funding from the European Research Council under the 
European Union’s Horizon 2020 research and innovation program, grant agreement 
number STG–677576 (“HARVEST”).

# Data Availability Statement {-}

All scripts and data used in the analysis are available on OSF 
(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
following the format provided by the rrtools package [@R-rrtools].
More detailed protocols are available on OSF and protocols.io.
